The role of prenatal conditions on brain mitochondrial physiology and prey detection in a lizard

### 2025-03-16 ms.qmd
updates

---
docx: 
reference-doc: "./bib/tmpl.docx"
csl: "./bib/proceedings-of-the-royal-society-b.csl"
bibliography: "./bib/refs.bib"
execute:
  echo: false
  error: false
  cache: false
  warning: false
crossref:  
  fig-title: Figure    # (default is "Figure")
  title-delim: —     # (default is ":")
  fig-prefix: Fig.   # (default is "Figure")
  tbl-prefix: Tab.   # (default is "Table")
editor_options: 
  chunk_output_type: console
---

@@

@@

@@

Cognition encompasses how animals acquire, process, and store information, enabling perception, learning, memory, and decision-making [@shettleworth]. It is essential for survival and reproduction, allowing individuals to adapt to their environment effectively [@dukas_evolutionary_2004]. However, cognitive abilities can vary considerably between individuals, with differences arising from genetic factors, environmental conditions, or a combination of both [@sakata_neural_2000; @dukas_evolutionary_2004]. The prenatal environment, in particular, plays a critical role in shaping brain development and cognition across a wide range of species [@sakata_neural_2000; @zhu_prenatal_2004; @crino_corticosterone_2014-learn; @amiel_effects_2017]. For instance, plenty of studies have investigated how prenatal conditions can influence learning abilities and memory across different taxa [@zhu_prenatal_2004; @crino_corticosterone_2014-learn; @bebus_associative_2016; @abayarathna_effects_2020]. However, the impact of early-life environments on other cognitive domains has been less studied [but see @burger_antipredator_1998; @vila_pouca_quantity_2019]. A fundamental cognitive skill is acquiring information from the environment [@shettleworth]. Perception of different stimuli is essential for locating food, avoiding predators, or interacting with conspecifics [@burger_antipredator_1998; @desfilis2003stimulus; @recio2023conspecific]. Variations in the ability to detect or recognize some relevant stimuli can have significant consequences for an individual’s fitness. For example, individuals better at detecting prey might be more competitive in foraging contexts, which will affect the energy they can obtain and allocate for other processes like growth or reproduction. Therefore, understanding how prenatal conditions affect the ability to detect prey can be essential to understanding the broader consequences of early-life environments on fitness. Yet, little is known about how prenatal conditions influence prey-related cognition [but see @burger_effects_1990].

The prenatal environment influences cognitive abilities because the brain is particularly sensitive to environmental inputs during the early stages of development [@zhu_prenatal_2004]. Early-life conditions can shape cognition by altering gene expression [@zhou2020effects], neurotransmitter production [@amani2021perinatal], or brain structure [@amiel_effects_2017]. One key mechanism linking cognition and brain function is mitochondrial physiology and their effects on energy production and oxidative stress [@siegel1994basic; @du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018]. Given the high energetic demands of cognitive processes [@mcnay_decreases_2000; @mann_coupling_2021; @alexandrov_neuronal_2022], cognitive performance is expected to depend on efficient mitochondrial respiration. Additionally, high cognitive abilities are typically associated with increased neuron density and functionality [@lefebvre_taxonomic_2011; @amiel_effects_2017], both of which can be impaired by excessive ROS production and oxidative stress byproduct of mitochondrial function [@finkel_oxidants_2000; @zhu_prenatal_2004; @du_dynamic_2009; @gong_chronic_2011; @hoffmann_mitochondrion_2018]. For instance, @hara_presynaptic_2014 found that performance in visuospatial working-memory tasks correlated negatively with the number of round-shaped mitochondria — associated with higher oxidative stress — per presynaptic bouton. These findings emphasize the role of mitochondrial action in shaping cognitive abilities. However, how the early environment influences mitochondrial physiology and how that translates into cognitive consequences remains largely unknown in non-mammalian animals [see @chaudhari2022early for a review]. In particular, whether and how these interactions affect perception abilities, such as prey detection, is still largely unexplored.

@@

In ectotherms, thermal conditions can be a significant source of maternal stress. When animals experience temperatures outside their optimal thermal range, they activate the physiological stress response, leading to increased GC production that can be passed to the offspring [see @Crino_2023]. However, the temperature can also directly affect offspring development beyond maternal effects, particularly during early life [@noble_developmental_2018; @crino2024eggs]. The prenatal thermal environment can play a critical role in shaping mitochondrial function, altering energy metabolism or oxidative stress [@stier2022experimental; @crino2024eggs]. For example, Red-eared slider turtle (_Trachemys scripta elegans_) hatchlings incubated at 26.5 ºC exhibited greater oxidative damage compared to those incubated at temperatures 0.7 to 1.2 ºC warmer [@treidel2016temperature]. Therefore, the combined effects of prenatal GCs and temperature can have profound consequences for mitochondrial function, with significant repercussions for brain development and cognition. Yet, we have little evidence of how the interactive effects of prenatal GCs and temperature can impact cognitive abilities through mitochondrial physiology.

Here, we investigated the combined effects of prenatal temperature and corticosterone (CORT) - the main GC in reptiles - on mitochondrial functioning and prey detection in the delicate skink (_Lampropholis delicata_). We manipulated prenatal CORT levels and incubation temperature, then assessed hatchlings’ ability to detect chemical and visual prey stimuli through behavioral tests. After these tests, we examined mitochondrial function in different brain regions associated with processing chemical and visual cues. We hypothesized that prenatal CORT and temperature would interact to influence mitochondrial physiology and prey detection. GCs can alter mitochondrial function by increasing oxidative stress [reviewed in @costantini2011meta] and reducing mitochondrial efficiency [@gong_chronic_2011;  but see @crino2024eggs]. Similarly, high temperatures can decrease mitochondrial efficiency but also oxidative damage [@treidel2016temperature; @crino2024eggs]. Therefore, we predicted that prenatal CORT exposure would impair mitochondrial function by increasing ROS levels and reducing energy availability, disrupting cognitive performance. However, higher incubation temperatures would increase performance only when the cost of higher oxidative stress overcome the benefits of increased mitochondrial efficiency. In addition, we predicted complex interactions between CORT and incubation temperature, as both factors negatively impact mitochondrial efficiency but have opposite effects on oxidative stress. Finally, since early environmental effects on mitochondrial function can vary across brain regions [@coomber_independent_1997], we expected regional differences in mitochondrial physiology accross treatments, which in turn would lead to stimulus-dependent variations in detection abilities. Our study provides insights into how prenatal conditions shape mitochondrial physiology and cognitive abilities, emphasizing the importance of early-life environments in cognitive development.

@@

#### Husbandry

@@

*Hatchlings* -- Incubators were checked three times a week for hatchlings. We measured and weighted lizards immediately after hatching. We employed a rule to obtain Snout-vent length (SVL) and Tail length (TL) to the nearest mm, and a digital scale (± 0.001g error) to measure mass. We then placed hatchlings in individual enclosures (18.7L x 13.2W x 6.3H cm) and provided them with nonstick matting and a small water dish until the beginning of the experiment. During this period, lizards were given water dayly and received 3-6 small *A. domestica* crickets three times a week. All care otherwise follows similar protocols to adults (see above).

Two weeks before we started the tests (see below), lizards were moved to the experimental arena (see @fig-Methods) for acclimatization. The arenas were individual medium size (41 L x 29.7 W x 22 H cm) plastic containers with a shelter (9 L x 6 W x 1.5 H cm) on one of the extremes and a water dish on the middle of the arena. These arenas were placed in two rooms in six racks, each with its own CCTV system (device model DVR-HP210475) that allowed us to record lizards' behaviour during the experiment (see details below). The number of lizards per treatment in each rack was counterbalanced to control for any effect of the room or the position of the lizard in the rack. During acclimatization, lizards were fed with only one cricket per day dusted with calcium and multivitamin, and water was supplied *ad libitum*. We provided a temperature gradient by means of a heat cord and heat lamps in a 12 h light: 12 h dark cycle. The rooms temperature was set to between 22-24 Celsius. After the tests, animals were euthanized and mitochondrial physiology was analyzed in various brain regions (see Brain mitochondrial activity protocol below).

#### CORT and Temperature manipulation
To test the interactive effects of CORT and incubation temperature, we manipulated CORT concentrations in eggs and incubated them under one of two temperature regimes (Cold - 23 ± 3 ºC or Hot - 28 ± 3 ºC) in a 2x2 factorial design (@fig-Methods A). Eggs were topically supplied with: a) 5 µL of crystalline corticosterone (Sigma, Cat. No. C2505) dissolved in 100% ethanol at a final 10 pg CORT/mL concentration (CORT treatment), or b) an equal volume of 100% Ethanol (Control treatment). We selected doses based on previous studies where CORT treatment increased mean yolk CORT levels ~3.7x higher than control eggs [@crino2024eggs]. Then, eggs were incubated in one of the two previously mentioned temperature regimes. These temperatures are within the natural limits in _L. delicata_ [@cheetham2011embryonic].

The number of eggs per clutch assigned to each hormone and temperature treatment was counterbalanced in a partial split-clutch design. At least one egg per clutch was assigned to each treatment, while in cluthces bigger than four the remaining eggs were randomly assigned to one of the treatments. When we found less than four eggs in a clutch, we assigned each egg randomly to one of the treatments.

#### Prey discrimination tests
The experiment involved presenting lizards with various stimuli from different prey types, then recording and analyzing their behaviour towards each stimulus. We followed the methods employed by @desfilis2003stimulus and @recio_prey_2021, with some modifications relevant to the objectives of this study. Lizards were presented chemical or visual stimuli from a familiar and unfamiliar prey, with a total of four trials per individual (@fig-Methods C).

We used crickets (_A. domestica_) as the familiar prey and mealworm larvae (_Tenebrio molitor_) as the unfamiliar prey. None of the species are included in the natural diet of the lizards [@lunney1989diets], but both are reported to be eaten in lab conditions [@michelangeli2016sa]. We expected to see differences between known and unknown prey because previous experience may influence stimuli perception through habituation or sensitisation [@burger_effects_1990; @burger_effects_1991; @desfilis2003stimulus], and we included familiarity with the prey as a factor in our analyses (see below).

Each stimulus was presented inside a transparent plastic vessel containing a white, two-chambered device (see @fig-Methods B) made of polylactic acid. In chemical trials, the prey was placed in the closed chamber at the back of the device, making it invisible to the lizard, while in visual trials, the prey was placed in the front chamber. Holes in both the device and the front sides of the transparent vessel (see @fig-Methods B, C) allowed chemical cues to be released; however, these holes were sealed with silicone in the visual trials. To increase the availability of chemical cues, we glued a piece of filter paper (left for at least 8 hours in one of the prey's enclosures: _A. domestica_ or _T. molitor_) to the device during chemical trials. In visual trials, the filter paper was placed in an empty box for the same duration under identical conditions. In both chemical and visual trials, the prey remained inside the vessel to control for potential acoustic cues. The order of stimulus presentation was counterbalanced across treatments.

Each trial began by placing the experimental device in the side of the arena opposite to the shelter (see arena in @fig-Methods A), and then removing the shelter. The water cup had already been removed. We recorded the lizard's behaviour for approximately one hour. We assessed lizards' ability to detect each stimulus ('Detection latency') by recording the time from when the lizard resumed normal activity (i.e., walking for at least 5 consecutive seconds; T~0~ in @fig-Methods D) until the first interaction with the stimulus (T~D~ in @fig-Methods D). We consider an interatcion when the lizard touched the front of the vessel or the filter paper with its snout for more than five consecutive seconds.

To control for potential differences in hunger levels, all lizards fasted for two days before the experiment, a period considered harmless for this species [@young2022physiological]. Besides, after each trial, the lizards were given a cricket to assess their motivation to forage. The cricket was left in the enclosure for one hour, and we recorded whether the lizard ate it (recorded as 1) or not (recorded as 0). In `r sum(is.na(clean_df$motivation))` videos, the camera stopped recording before the end of the motivation test (T~f~ in @fig-Methods D), so motivation was recorded as NA. We used their performance in the motivation test as a covariate in the analyses (see below).

@@

#### Brain mitochondrial activity
Two months after the completion of the tests, we euthanized lizards using an injectable anaesthetic followed by decapitation. We injected peritoneally 10 mg/kg of a 10 mg/mL alfaxan solution and then, after several minutes, we evaluated the lizard's response by turning it upside down, and by testing the pinching relfex in one of the front limbs. Lizards were decapitated with surgical scissors once they were not responsive. This protocol was approved by the Animal Ethics Committee of the Australian National University (Protocol number: A2022_33). After decapitation, the head was opened and the brain was dissected. We extracted manually two main regions of the brain, the olfactory bulbs and the optic tecta, as they are associated with lizards' chemical and visual perception, respectively [@wyneken2007reptilian]. Both regions were transferred immediately to 1.5mL centrifuge tubes containing 100µL of 1X PBS.

Both regions were transferred immediately to 1.5 mL centrifuge tubes containing 100 µL of cold 1X PBS and kept on ice until further processing. For the olfactory bulbs, the tissue was mechanically homogenized by placing the tissue in the well of a 100 µm mesh filter (pluriStrainer) affixed atop a 1.5 mL centrifuge tube, then mashed with the rubber end of an insulin syringe stopper. The resulting olfactory bulb homogenate was then rinsed through the filter with 1 mL of cold 1X PBS to prepare a homogenate suspension. Optic tecta were mechanically homogenized the same way as olfactory bulbs, but were first enzymatically digested by incubating the tissue in 100µL of 125 U/mL collagenase (type II) for 30-min at 32°C. We enzymatically digested optic tecta but not olfactory bulbs because in a pilot study, enzymatic digestion prior to mechanical digestion improved the yield of viable cells in optic tecta homogenate, but did not noticeably affect the yield of viable cells in olfactory bulbs homogenate (unpublished data). Following homogenization, we centrifuged each sample at 1000 RCF for 10 minutes to pellet cells, then removed the supernatant (hereafter, this process referred to as ‘washing’) and resuspended the cells in 500 µL 1x PBS. This step was to performed to remove cellular debris from homogenates. From each 500 µL suspension of homogenate collected on a given trial day, we first added 100 µL of homogenate to a pooled sample of each tissue type to use for single-color controls, and the remaining 40 0µL of homogenate was split among two 200 µL aliquots. One aliquot was used fresh to measure mitochondrial function (mitochondrial density, membrane potential, ROS), one aliquot was cryopreserved for later measurements of oxidative damage (8-OHdG, lipid peroxidation), and the third aliquot was cryopreserved for a different experiment. Aliquots were cryopreserved in 1 mL solution of 1% Neutral-Buffered Formalin (as a fixative agent), 1X Tris-EDTA (chelates metals that can damage DNA during freezing) and 10% DMSO (a cryoprotectant), then stored at -20 °C until oxidative damage assays.

#### Statistical analyses

We performed the analyses for each brain region/stimulus and each variable (mitochondrial density, mitochondrial potential, ROS, DNA damage, lipid peroxidation, and detection latency) separately. We fitted first a set of preliminary models where we included the main effects of hormone (CORT versus Control), temperature (Cold versus Hot), and their interaction. We also added to these models the sex and age of the lizards at the time where the trials started (for detection latency), or when the lizards were euthanized (for all mitochondrial-related variables). For detection latency we also included as fixed factors lizards' previous experience with the prey (familiar versus unknown), their performance on the motivation test [if they ate the cricket (1), or not (0)], and the interaction between CORT and motivation, as CORT can impact apetite [@conde2018stress]. We also included for all the models the clutch identity as a random factor. Since detection latency was fitted for both types of prey in a repeated measures design, these models also included individual identity as a random factor. Structure and results of these models are provided on Tables S1-S12 in Supplementary Material. After the preliminary models, we fitted a set of definitive models where those factors that were not significant in the preliminary models were excluded except the main interest factors: CORT, temperature and their interaction. Random factors remain the same in the definitive models. All the response variables were mean centered and standardized by dividing by two times the standard deviation. Before standardization, mitochondrial density, DNA damage, lipid peroxidation, and detection latency were log-transformed.

All the models were run using the brm function from the _brms_ package [@burkner2017brms] in R (version 4.4.0) [@R]. We ran four parallel MCMC chains of 8000 iterations for each model, with a warmup period of 2000 iterations. The error structure was modelled assuming a gaussian distribution for all variables. To test for differences between treatments, we made contrasts between treatments using the posterior distributions of the parameters. We employed pMCMC values to tests if the contrasts were different from zero [@endo2019introduction]. We considered an effect statistically significant if pMCMC < 0.05.

To further explore the relationships between mitochondrial physiology and detection latency, we followed Structural Equations Modelling approach. We fitted a multivariate model using the _brms_ package [@burkner2017brms] in R (version 4.4.0) [@R] for each brain region/stimulus separately. We included in the model all the variables of interest and their interactions structured following a specific hypothesis (see @fig-sem_results_OB or @fig-sem_results_OT). We also included for each variable those factors that were significant in our definitive models but we excluded the non-significant ones. Because experience with prey did not affect lizards' behaviour (see Tables S11, S12), and to reduce the complexity of the models, we averaged lizards detection latency across both types of prey and excluded lizard identity from the random factors. Clutch identity was included as a random factor for all variables. We ran four parallel MCMC chains of 8000 iterations for each model, with a warmup period of 2000 iterations. The error structure was modelled assuming a gaussian distribution. Rescor was set as true to control for potential correlations between the residuals of the variables.

>>>>

---
title: "The role of prenatal conditions on brain mitochondrial physiology and prey detection in a lizard"
docx: 
reference-doc: "./bib/tmpl.docx"
csl: "./bib/proceedings-of-the-royal-society-b.csl"
bibliography: "./bib/refs.bib"
execute:
  echo: false
  error: false
  cache: false
  warning: false
crossref:  
  fig-title: Figure    # (default is "Figure")
  title-delim: —     # (default is ":")
  fig-prefix: Fig.   # (default is "Figure")
  tbl-prefix: Tab.   # (default is "Table")
editor_options: 
  chunk_output_type: console
---

@@

Pablo Recio^1^,$\ddagger$ , Dalton C. Leibold^1^, Ondi L. Crino ^1,2^, Christopher R. Friesen^3^, Daniel W.A. Noble^1^

^1^ Division of Ecology and Evolution, Research School of Biology, The Australian National University, Canberra, ACT 2601, Australia  
^2^ Flinder's University, College of Science and Engineering, Bedford Park, SA 5042, Australia  
^3^ University of Wollongong, Wollongong, NSW 2500, Australia  
$\ddagger$ Corresponding author: pablo.reciosantiago@anu.edu.au

ORCID:

Pablo Recio ORCID: 0000-0002-5890-0218  
Dalton C. Leibold ORCID: 0000-0001-9645-2033  
Ondi L. Crino ORCID: 0000-0001-5700-1387  
Christopher R. Friesen ORCID: 0000-0001-5338-7454  
Daniel W.A. Noble ORCID: 0000-0001-9460-8743

@@

## Abstract
Cognitive processes such as the ability to perceive prey, are crucial for survival and reproduction. However, early-life conditions like stress-related hormones or thermal environments, can shape cognitive abilities by influencing brain structure and function. Mitochondrial physiology is a central mechanism underlying these effects. Here, we investigated the combined influence of prenatal corticosterone (CORT) — a key stress hormone in reptiles — and incubation temperature on mitochondrial function and prey detection in the delicate skink (_Lampropholis delicata_). We manipulated egg CORT levels and incubation temperature, then assessed hatchlings’ ability to detect chemical and visual prey stimuli. Using flow cytometry, we measured mitochondrial function, reactive oxygen species (ROS) production, and oxidative stress in the olfactory bulbs and optic tecta — two brain regions involved in sensory processing. We predicted that CORT exposure and high temperatures would interact to shape perception of prey via changes in mitochondrial respiration and oxidative damage. While mitochondrial function remained robust to early conditions, CORT and temperature interacted to influence oxidative stress in the optic tecta but not the olfactory bulbs. Additionally, CORT-treated lizards responded faster when presented with chemical cues but not visual stimuli. However, mitochondrial function and oxidative stress were not directly related to prey detection performance. These findings suggest that prenatal conditions can have lasting effects on brain physiology and cognition, though the role of mitochondrial function in these processes remains unclear. Future research should explore alternative mechanisms linking early-life stress, metabolic function, and cognitive performance.

@@

Cognition encompasses how animals acquire, process, and store information, enabling perception, learning, memory, and decision-making [@shettleworth]. It is essential for survival and reproduction, allowing individuals to adapt to their environment effectively [@dukas_evolutionary_2004]. However, cognitive abilities can vary considerably between individuals, with differences arising from genetic factors, environmental conditions, or a combination of both [@sakata_neural_2000; @dukas_evolutionary_2004]. The prenatal environment, in particular, plays a critical role in shaping brain development and cognition across a wide range of species [@sakata_neural_2000; @zhu_prenatal_2004; @crino_corticosterone_2014-learn; @amiel_effects_2017]. For instance,  studies have identified how prenatal conditions can influence learning abilities and memory across different taxa [@zhu_prenatal_2004; @crino_corticosterone_2014-learn; @bebus_associative_2016; @abayarathna_effects_2020]. However, the impact of early-life environments on other cognitive domains has been poorly studied [but see @burger_antipredator_1998; @vila_pouca_quantity_2019]. Perception is an often overlooked, but fundamental skill [@shettleworth]. Perceiving different stimuli is essential for locating food, avoiding predators, or interacting with conspecifics [@burger_antipredator_1998; @desfilis2003stimulus; @recio2023conspecific]. Variation in the ability to detect or recognize relevant stimuli can have significant consequences for an individual’s fitness [@pyke1984optimal]. For example, visual acuity in the bluegill sunfish (_Lepomis macrochirus_) allow animals to detect prey at larger distances, which can increase their foraging efficiency and energy intake [@hairston1982fish]. Therefore, understanding how prenatal conditions affect the ability to detect prey can be essential to understanding the broader consequences of early-life environments on fitness. Yet, little is known about how prenatal conditions influence prey-related cognition [but see @burger_effects_1990].

Prenatal environments influence cognitive abilities because the brain is particularly sensitive to environmental inputs during early stages of development [@zhu_prenatal_2004]. Early-life conditions can shape cognition by altering gene expression [@zhou2020effects], neurotransmitter production [@amani2021perinatal], or brain structure and function [@amiel_effects_2017]. Long-lasting effects on energy production and oxidative stress in the brain may be a key mechanism by which cognitive function is impacted by developmental environments  [@siegel1994basic; @du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018]. Given the high energetic demands of cognitive processes [@mcnay_decreases_2000; @mann_coupling_2021; @alexandrov_neuronal_2022], cognitive performance is expected to depend on efficient mitochondrial respiration. Additionally, high cognitive abilities are typically associated with increased neuron density and functionality [@lefebvre_taxonomic_2011; @amiel_effects_2017], both of which can be impaired by excessive reactive oxygen species (ROS) production and oxidative stress [@finkel_oxidants_2000; @zhu_prenatal_2004; @du_dynamic_2009; @gong_chronic_2011; @hoffmann_mitochondrion_2018]. For instance, Hara et al. (2014) found that performance in visuospatial working-memory tasks correlated negatively with the number of round-shaped mitochondria — associated with higher oxidative stress — per presynaptic terminal. These findings emphasize the role of mitochondrial action in shaping cognitive abilities. However, how the early environment influences mitochondrial physiology and how it may translate into cognitive consequences remains largely unknown in non-mammalian animals [see @chaudhari2022early for a review]. In particular, whether and how these interactions affect perceptual abilities, such as prey detection, is still largely unexplored.

@@

In ectotherms, thermal stress experienced by mothers can increase maternal circulating glucocorticoids. Activation of the physiological stress response and higher GC concentrations can subsequently impact offspring [see @Crino_2023]. However, temperature is also known to directly affect offspring development beyond maternal effects, particularly during early life [@noble_developmental_2018; @crino2024eggs]. The prenatal thermal environment can play a critical role in shaping mitochondrial function, altering energy metabolism or oxidative stress [@stier2022experimental; @crino2024eggs]. For example, Red-eared slider turtle (_Trachemys scripta elegans_) hatchlings incubated at 26.5 ºC exhibited greater oxidative damage compared to those incubated at temperatures 0.7 to 1.2 ºC warmer [@treidel2016temperature]. Therefore, the combined effects of prenatal GCs and temperature can have profound consequences for metabolic function, with significant repercussions for brain development and cognition. Yet, we have little evidence of how the interactive effects of prenatal GCs and temperature can impact cognitive abilities by developmental programming of metabolic pathways.

Here, we investigated the combined effects of prenatal temperature and corticosterone (CORT) - the main GC in reptiles - on mitochondrial function and prey detection in the delicate skink (_Lampropholis delicata_). We manipulated eggs' CORT levels and incubation temperature, then assessed hatchlings’ prey detection ability and mitochondrial function in brain regions related to processing chemical and visual cues. We hypothesized that CORT would impair prey detection through mitochondrial pysiology by decreasing respiration efficiency [@gong_chronic_2011;  but see @crino2024eggs] and increasing oxidative stress [reviewed in @costantini2011meta]. In contrast, we predicted that low incubation temperatures would only enhance beahvioural performance if the benefits of mitochondrial efficiency outweigh the costs of higher oxidative stress [@treidel2016temperature; @crino2024eggs]. We also expected complex interactions between CORT and temperature, as they have opposing effects on mitochondrial efficiency. Finally, we predicted region-dependet effects of early-life conditions on mitochondrial function [@coomber_independent_1997] that would lead to stimulus-specific differences in prey detection.

@@

@@

*Hatchlings* -- Incubators were checked three times a week for hatchlings. We measured and weighted lizards immediately after hatching. We used a ruler to measure snout-vent length (SVL) and tail length (TL) to the nearest mm, and a digital scale (± 0.001g error) to measure mass. We then placed hatchlings in individual enclosures (18.7L x 13.2W x 6.3H cm) and provided them with nonstick matting and a small water dish until the beginning of the experiment. During this period, lizards were given water daily and received 3-6 small *A. domestica* crickets three times a week. All care otherwise follows similar protocols to adults (see above).

#### CORT and Temperature manipulation

To test the interactive effects of CORT and incubation temperature, we manipulated CORT concentrations in eggs and incubated them under one of two temperature regimes (see @fig-Methods A). We used a partial split clutch design where eggs from a given clutch were distributed equally across the four treatments when clutch sizes were larger than four and randomly across treatments when less than four. Eggs were topically supplied with either: a) 5 µL of crystalline corticosterone (Sigma, Cat. No. C2505) dissolved in 100% ethanol at a final 10 pg CORT/mL concentration (CORT treatment), or b) an equal volume of 100% Ethanol (Control treatment). We selected doses based on our previous study where CORT treatment increased mean yolk CORT levels ~3.7x higher than control eggs [@crino2024eggs]. Eggs were then incubated at either cold  (23 ± 3 ºC) or hot (28 ± 3 ºC) incubation temperatures in a 2x2 factorial design (Fig. 1 A). These temperatures are within the natural limits in _L. delicata_ [@cheetham2011embryonic].

#### Prey discrimination tests
Two weeks before we started the tests (see below), lizards were moved to the experimental arena (see Fig. 1) for acclimatization. The arenas were individual medium size (41 L x 29.7 W x 22 H cm) plastic containers with a shelter (9 L x 6 W x 1.5 H cm) on one of the extremes and a water dish on the middle of the arena. These arenas were placed in two rooms in six racks, each with its own CCTV system (device model DVR-HP210475) that allowed us to record lizards’ behaviour during the experiment (see details below). The number of lizards per treatment in each rack was counterbalanced to control for any effect of the room or the position of the lizard in the rack. During acclimatization, lizards were fed with only one cricket per day dusted with calcium and multivitamin, and water was supplied ad libitum. We provided a temperature gradient by means of a heat cord and heat lamps in a 12 h light: 12 h dark cycle. The room temperature was set to between 22-24 Celsius. After the tests, animals were euthanized and metabolic function was analyzed in various brain regions (see Brain mitochondrial activity protocol below).

The experiment involved presenting lizards with various stimuli from different prey types, then recording and analyzing their behaviour towards each stimulus. We followed the methods employed by @desfilis2003stimulus and @recio_prey_2021, with some modifications relevant to the objectives of this study. Lizards were presented chemical or visual stimuli from a familiar and unfamiliar prey, with a total of four trials per individual (@fig-Methods C).

We used crickets (_A. domestica_) as the familiar prey and mealworm larvae (_Tenebrio molitor_) as the unfamiliar prey. We expected to see differences between known and unknown prey because previous experience may influence stimuli perception through habituation or sensitisation [@burger_effects_1990; @burger_effects_1991; @desfilis2003stimulus], and we included familiarity with the prey as a factor in our analyses (see below).

Each stimulus was presented inside a transparent plastic vessel containing a white, two-chambered device (see @fig-Methods B) made of polylactic acid (PLA). In chemical trials, the prey was placed in the closed chamber at the back of the device, making it invisible to the lizard, while in visual trials, the prey was placed in the front chamber. Holes in both the device and the front sides of the transparent vessel (see @fig-Methods B, C) allowed chemical cues to be released; however, these holes were sealed with silicone in the visual trials. To increase the availability of chemical cues, we glued a piece of filter paper (left for at least 8 hours in one of the prey's enclosures: _A. domestica_ or _T. molitor_) to the device during chemical trials. In visual trials, the filter paper was placed in an empty box for the same duration under identical conditions. In both chemical and visual trials, the prey remained inside the vessel to control for potential acoustic cues. The order of stimulus presentation was counterbalanced across treatments.

Each trial began by placing the experimental device at the side of the arena opposite to the shelter (see arena in @fig-Methods A), and then removing the shelter. The water cup had already been removed. We recorded the lizard's behaviour for approximately one hour. We assessed lizards' ability to detect each stimulus ('Detection latency') by recording the time from when the lizard resumed normal activity (i.e., walking for at least 5 consecutive seconds; T~0~ in @fig-Methods D) until the first interaction with the stimulus (T~D~ in @fig-Methods D). Lizards were considered to have interacted with the object when the lizard touched the front of the vessel or the filter paper with its snout for more than five consecutive seconds.

To control for potential differences in hunger levels, all lizards were fasted for two days before the experiment, a period considered harmless for this species [@young2022physiological]. After each trial, the lizards were also given a cricket to assess their motivation to forage. The cricket was left in the enclosure for one hour, and we recorded whether the lizard ate it (recorded as 1) or not (recorded as 0). In `r sum(is.na(clean_df$motivation))` videos, the camera stopped recording before the end of the motivation test (T~f~ in @fig-Methods D), so motivation was recorded as NA. We used their performance in the motivation test as a covariate in the analyses (see below).

@@

#### Brain metabolic function

_Brain dissection and homogenization_
Two months after the completion of the tests, we euthanized lizards using an injectable anaesthetic followed by decapitation. We injected peritoneally 10 mg/kg of a 10 mg/mL alfaxan solution and then, after several minutes, we evaluated the lizard’s righting response and pinching reflex in one of the front limbs. Lizards without responses were decapitated with surgical scissors. This protocol was approved by the Animal Ethics Committee of the Australian National University (Protocol number: A2022_33). After decapitation, the head was opened and the brain was dissected. We extracted manually two main regions of the brain, the olfactory bulbs and the optic tecta, as they are associated with chemical and visual perception in lizards [@wyneken2007reptilian]. Both regions were transferred immediately to 1.5mL centrifuge tubes containing 100µL of 1X phosphate buffered saline (PBS).

Tissue suspected in PBS was then homogenized. For the olfactory bulbs, the tissue was mechanically homogenized by placing the tissue in the well of a 100 µm mesh filter (pluriStrainer) affixed atop a 1.5 mL centrifuge tube, then mashed with the rubber end of an insulin syringe stopper. The resulting olfactory bulb homogenate was then rinsed through the filter with 1 mL of cold 1X. Optic tecta were mechanically homogenized the same way as olfactory bulbs, but were first enzymatically digested by incubating the tissue in 100µL of 125 U/mL collagenase (type II).

Homogenates were split among two aliquots: one was used fresh to measure mitochondrial density, membrane potential and ROS, while the other one was cryopreserved for later measurements of DNA damage and lipid peroxidation. Cryopreservation was made by suspending the homogenates in 1 mL solution of 1% Neutral-Buffered Formalin, 1X Tris-EDTA, and 10% DMSO, then stored at -20 °C until oxidative damage assays.

_Metabolic function_
Fresh homogenate suspensions were stained with 5 µL of a fluorescent probe mix containing equal parts 5 µM MitoTracker Deep Red FM, 2.5 µM MitoTracker Orange CMTMRos, and 50 µM MitoSOX Red. We used these fluorescent probes as indicators of mitochondrial density, mitochondrial membrane potential, and superoxide (ROS) production, respectively. We also added 5 µL of #µM Hoechst 33342 Nuclear Viability Dye to each sample, which we used to distinguish live, viable, intact cells from cellular debris. Within two hours after dissection, these samples were analyzed by flow cytometry.

_Oxidative Damage_
Assays of oxidative damage from cryopreserved samples were performed within 6 months of the initial processing and analysis of fresh samples. On the day of oxidative damage assays, we rapidly thawed frozen samples, removed the cryopreservation solution (see above), and resuspended samples in 1 mL warm 1X Tris-EDTA. Afterward, the Tris-EDTA was removed, and samples were resuspended in 200 µL of warm 1X PBS containing 20 µL of #µM Hoechst 33342 Nuclear Viability Dye (for cell viability) and 20 µL of 100 µM BODIPY 665/676 Lipid Peroxidation Sensor (to measure lipid peroxidation). Following staining, we permeabilized the cell membranes incubating the samples in 200 µL warm 1X PBS containing 5 mM (6.14 mg/mL) digitonin. After permeabilization, we stained the samples with 20 µL of 70 µM 8-OHdG Polyclonal Antibody to measure oxidative damage on DNA. We left the homogenate overnight (~12 hours). The following day we counterstained the cells with 20 µL of #µM (100 µg/mL) H+G Goat Anti-Rabbit Conjugate Antibody with Alexa-Fluor 488 and analyze the samples in the flow cytometer.

_Flow Cytometry_
Flow cytometry assays were performed using a Becton Dickson LSRFortessa X-20 flow cytometer with the default wavelength filters on detectors. The detectors and voltage settings used in data acquisition for each were determined during pilot trials and kept consistent throughout the experiment. Data was imported into FlowJo (v. 10.1) for processing. We obtained the mean fluorescent intensity for mitochondrial density, mitochondrial potential, ROS production, DNA damage, and lipid peroxidation. For further details on the homogenization, staining, or flow cytometry assays, see Methods: flow cytometry in Supplementary Material

#### Statistical analyses
We performed the analyses for each brain region/stimulus and each variable (mitochondrial density, mitochondrial potential, ROS, DNA damage, lipid peroxidation, and detection latency) separately. We first fitted a set of preliminary models for all the interest varaibles (detection latency, mitochondrial density, mitochondrial potential, ROS, DNA damage, and lipid peroxidation) where we included the main effects of hormone (CORT versus Control), temperature (Cold versus Hot), and their interaction . Models also included the sex and age of the lizards at the time where the trials started (for detection latency) or when the lizards were euthanized (for all mitochondrial-related variables). For detection latency, we also included previous experience with the prey (familiar versus unknown), their performance on the motivation test [if they ate the cricket (1), or not (0)], and the interaction between CORT and motivation as fixed effects. We included the interaction between CORT and motivation because CORT can impact appetite [@conde2018stress]. All models included clutch identity as a random factor to account for clutch effects. For models of detection latency we also included a random effect of lizard identity because of our repeated measures design. The structure and results of these models are provided on Tables S1-S12 of the Supplementary Material. After the preliminary models, we simplified models by re-fitting to exclude factors that were not significant. However, our models always included the main interest factors: CORT, temperature and their interaction. Random factors remain the same for the final models. All the response variables were mean centered and standardized by dividing by two times the standard deviation [@gelman2008scaling]. Before standardization, mitochondrial density, DNA damage, lipid peroxidation, and detection latency were log-transformed.

All models were fit using using brm models from stan [@stan] in R (version 4.4. 0) [@R]. The error structure was modelled assuming a gaussian distribution for all variables. We ran four parallel MCMC chains of 8000 iterations for each model, with a warmup period of 2000 iterations. To test for differences between treatments, we made contrasts between treatments using the posterior distribution of relevant parameters. Significance was assessed by calculating the probability distribution parameter estimates and contrasts to test whether were they different from zero [@endo2019introduction]. We considered an effect statistically significant if pMCMC < 0.05.
